An interdisciplinary examination of stress and injury occurrence in athletes

### 2020-08-14 manuscript.tex
for mac

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\makeatother
\shorttitle{Stress and sports injury}
\author{Harry Fisher\textsuperscript{1}, Dr Marianne Gittoes\textsuperscript{1}, Professor Lynne Evans\textsuperscript{1}, Miss Leah Bitchell\textsuperscript{1}, Dr Richard Mullen\textsuperscript{2}, \& Dr Marco Scutari\textsuperscript{3}}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Cardiff Metropolitan University, Cardiff, United Kingdom\\\textsuperscript{2} Brunel University, London, United Kingdom\\\textsuperscript{3} Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA), Manno, Switzerland}
\authornote{12961 words in text body and refence section

Number of tables: 11

Number of figures: 10

@@

\begin{figure}[!h]

{\centering \includegraphics[width=5.67in]{figures_doc/Fig1}

\caption{Stress and injury model (Williams and Andersen, 1998).}\label{fig:fig1}
\end{figure}

@@

Postural stability was assessed with a modified version of the balance error scoring system (mBESS) based on the protocol recommended by (55).
In total, each trial of the mBESS was performed without shoes (56) and included six stances in the following order; dominant leg (DL; standing on the dominant foot with the non-dominant foot at approximately 30-degrees of hip flexion and 45-degrees of knee flexion),
non-dominant leg (NDL; standing on the non-dominant foot with the dominant foot at approximately 30-degrees of hip flexion and 45-degrees of knee flexion)
and tandem leg stance (TS; standing heel-to-toe with the non-dominant foot behind the dominant) on firm and foam (Alcan airex AG, Sins, Switzerland) surfaces respectively (Fig \ref{fig:fig2}).
To determine leg dominance, participants were asked their preferred leg to kick a ball to a target, and the chosen limb was labelled as dominant (57).
Participants were asked to maintain each stance for a total of 20-seconds.
Participants hands were placed on hips at the level of the iliac crests.
A Sony DSC-RX10 video camera (Sony Europe Limited, Surrey, United Kingdom) was used to record each participants performance during the mBESS.

\begin{figure}[!h]

\caption{mBESS positions (A-F). Top row, firm surface. Bottom row, foam surface. Left column, dominant leg stance. Middle column, non-dominant leg stance. Right column, Tandem leg stance.}\label{fig:fig2}
\end{figure}

@@

At the start of the academic year (September), coaches of sports teams at a British University and local sports clubs were contacted and informed about the study. With the coaches' permission, the lead researcher attended training sessions to inform athletes about the overall purpose of the study and the requirements of participation. Athletes who met the participation criteria and volunteered to take part in the study were invited to attend scheduled testing sessions. A repeated measures prospective cohort design was used to assess athletes' major life events, stress-related physiological markers and injury status over two consecutive 12-month periods. Each participant was asked to attend a total of three data collections over a 12-month period, with each data collection separated by a four-month interval (Fig \ref{fig:fig3}). Participants provided informed consent before data collection commenced.

\begin{figure}
\includegraphics[width=1\linewidth]{figures_doc/Fig3} \caption{Study design. For each time point (T), each box contains the number of participants who completed the data collection (\textit{n}), the measures used for data collection and the approximate date of the data collection.}\label{fig:fig3}
\end{figure}

For the first three data collections (T1, T2 and T3), participants followed the same protocol in a specific order (Fig \ref{fig:fig4}).
To ensure all measures could be collected within a viable time-frame, participants were separated into two groups.
The first group completed all computer-based measures followed by all physical measurements, whereas the second group completed all physical measurements followed by computer-based measures.
Participants were randomly assigned to one of the two groups and remained in those groups across all time points.

\caption{Outline of the protocol for each data collection.}\label{fig:fig4}
\end{figure}

@@

\begin{figure}[H]

\caption{A simple discrete Bayesian network contain nodes, possible states of the nodes and the arcs connecting nodes.}\label{fig:fig5}
\end{figure}

@@

Learning the structure of the network is an important step in BN modelling. The structure of a network can be constructed using expert knowledge and/or data-driven algorithm techniques (e.g., search and score, such as hill climbing and gradient descent algorithms; Scutari \& Denis, 2014).
The learned structure can then be used for inference by querying the network\footnote{The term ``query'' in relation to Bayesian Networks stems from Pearl's expert systems theory (1988). A query can be submitted to an expert (in this case, the network is the expert) to get an opinion, the expert then updates the querier's beliefs accordingly. Widely used texts on Bayesian Network analysis (71) have widely adopted the terminology in favour of that used in traditional statistics.} and obtaining the posterior probabilities of a particular node for a given query.
The posterior distribution can be obtained by \(Pr(X|E,B) = Pr(X|E,G,\Theta)\), where the learned network \(B\) with structure \(G\) and parameters \(\Theta\), are investigated with new evidence \(E\) using the information in \(B\) (38).
In the example network presented in Fig \ref{fig:fig5}, new values assigned to each of the parent nodes (e.g., both set to ``Low'') could be used to investigate what effect the new information has on the state of the child node (conditional probability of a particular state of the child node).
In a more complex network containing many nodes, the outcome of a particular node can be assessed conditional on the states of any subset of nodes in the network.
BNs therefore provide a unique and versatile approach to modelling a set of variables to uncover dependency structures within the data.

@@

To obtain the network structure, several steps were taken to ensure that both a theoretically realistic network, and a network that was an appropriate fit to the collected data, was used for inference.
Prior knowledge about the network structure was included by providing a list of arcs that are always \emph{restricted} from being in the network (blacklist), and a list of arcs that are always \emph{included} in the network (whitelist).
Additionally, there are several scoring functions such as Bayesian Information Criteria (BIC) and Bayesian Dirichlet equivalent uniform (BDeu) that can be used to compare network structures with certain nodes and arcs included or excluded (38).
To account for the repeated measures design employed in this study and to maximise the use of the data, pairs of complete cases (e.g., participants who completed T1 + T2, and T2 + T3) were used in a two-time Bayesian network (2TBN) structure (79).
In the 2TBN, variables measured at T2 could depend on variables measured at T1 (e.g., T1 \(\rightarrow\) T2) and variables measured at T3 could depend on variables measured at T2 (e.g., T2 \(\rightarrow\) T3).
However, arcs were blacklisted between T2 \(\rightarrow\) T1 and T3 \(\rightarrow\) T2 to preserve the order in which data was collected.
Variables were separated into two groups; ``explanatory'', for variables that were fixed (e.g., gender), or ``independent'', for variables that were measured at each time point and could vary during the study.
Independent variable names were suffixed with \_1 for time point T, and \_2 for time point T+1 (e.g., T1\_1 \(\rightarrow\) T2\_2 and T2\_1 \(\rightarrow\) T3\_2).
Formatting the data in this way meant participants who completed T1 and T2, but did not complete T3, could still be included in the analysis.
Table \ref{tab:table2} provides an example of the formatted data and demonstrates that participants 1 and 3 have complete data, and therefore have two rows of data each representing variables from T1 \(\rightarrow\) T2 and T2 \(\rightarrow\) T3, respectively.
Participant 2 did not complete the final data collection at T3 and therefore only has one row of data representing the variables collected at T1 and T2.
In addition to the blacklisted arcs between T2 \(\rightarrow\) T1 and T3 \(\rightarrow\) T2, the direction of arcs was restricted between independent variables and explanatory variables (e.g., independent \(\rightarrow\) explanatory); however, arcs were not restricted between explanatory \(\rightarrow\) independent variables.
Finally, arc direction was restricted between specific nodes within the explanatory variables.
Arcs from clevel \(\rightarrow\) gender, nlebase \(\rightarrow\) gender and nlebase \(\rightarrow\) ind\_team were included in the blacklist, as arcs in these directions did not make logical sense.
All subsequent models used the same blacklist.

\caption{\label{tab:table2}Example of the data arrangement used for the network.}
\centering
\begin{tabular}[t]{c|c|c}
\hline
\textbf{Participant} & \textbf{X\_1} & \textbf{X\_2}\\
\hline
1 & T1 & T2\\
\hline
1 & T2 & T3\\
\hline
2 & T1 & T2\\
\hline
3 & T1 & T2\\
\hline
3 & T2 & T3\\
\hline
\end{tabular}
\end{table}

@@

The subscales representing the BAS (RR, RI, GDP and I) showed limited connection to other variables in the network.
Therefore, competing models were examined and BIC scores compared to establish the model with the best fit to the data (values are shown relative to the highest value).
RI provided the highest BIC value (-3,563.13), compared to RR (-3,579.10), GDP (-3,582.39) and I (-3,582.89).
Including all the variables (RR, RI, GDP and I) resulted in a significantly lower score -4,463.25) indicating that including all the variables was not beneficial to the model structure and did not offset the cost of the additional parameters. Therefore, only RI was included in the final structure.
Finally, both total score and asymmetry for balance were included in the initial network. However, visual inspection of the network revealed no arcs between bal\_asym\_1 or bal\_asym\_2 and any other node in the network.
Therefore, balance asymmetry was removed from the final network structure.
To summarise, Table \ref{tab:table3} reports the variables that were included in the final network structure.

@@

\caption{\label{tab:table3}Variables included in the final Bayesian network structure.}
\centering
\begin{tabular}[t]{l|>{\raggedright\arraybackslash}p{5.2cm}|c|c}
\hline
Variable & Definition & State 1 & State 2\\
\hline
clevel & Current competitive level & Club\_university\_county & National\_international\\
\hline
gender & Gender of the participant & Female & Male\\
\hline
hours & Number of hours spent training per week & 0-9 (Low) & >9-35 (High)\\
\hline
ind\_team & Participate in an individual or team based sport & Individual & Team\\
\hline
pi & Previous injury - Whether an injury had been sustained in the previous 12 months prior to the study & No Injury & Injury\\
\hline
nlebase & Untransformed NLE at TP 1 & 0-13 (Low) & >13-93 (High)\\
\hline
FFFS & Fight-Flight-Freeze System & 8-16 (Low) & >16-30 (High)\\
\hline
BIS & Behavioural Inhibition System & 17-38 (Low) & >38-68 (High)\\
\hline
RI & Reward Interest & 4-10 (Low) & >10-16 (High)\\
\hline
stiffness & Sum of all stiffness locations & 1543-2330 (Low) & >2330-4518 (High)\\
\hline
HRV & Root mean squared difference of successive RR intervals & 2.03-4.01 (Low) & >4.01-5.94 (High)\\
\hline
balance & Total balance score & 5-15 (Low) & >15-46 (High)\\
\hline
nlelg\_1 & Log NLE at TP 1 & 0-2.64 (Low) & >2.64-4.54 (High)\\
\hline
nlelg\_2 & Log NLE at TP 2 & 0-3.04 (Low) & >3.04-5.19 (High)\\
\hline
nlelg\_3 & Log NLE at TP 3 & 0-3.18 (Low) & >3.18-4.79 (High)\\
\hline
\end{tabular}
\end{table}

@@

During the study, 46\% (\emph{n} = 117) of participants reported at least one injury with an average severity of 11 \(\pm\) 31, days (range = 2 - 365 days).
Both male and female participants reported a greater number of acute compared to chronic injuries (male, acute = 85 {[}69\%{]}, chronic = 39 {[}31\%{]}; female, acute = 38 {[}72\%{]} chronic = 15 {[}28\%{]}), and non-contact injuries were more common than contact injuries (male, non-contact = 83 {[}67\%{]}, contact = 39 {[}31\%{]}; female, non-contact = 35 {[}66\%{]} contact = 18 {[}34\%{]}).
Table \ref{tab:table4} shows the number and percentage of injury types sustained by both male and female participants.

@@

\caption{\label{tab:table4}The number and percentage (\%) of types of injuries sustained by male and female participants.}
\centering
\begin{tabular}[t]{l|c|c|c|c}
\hline
\multicolumn{1}{c|}{\textbf{ }} & \multicolumn{2}{c|}{\textbf{Female}} & \multicolumn{2}{c}{\textbf{Male}} \\
\cline{2-3} \cline{4-5}
\textbf{} & \textbf{Lower body} & \textbf{Upper body} & \textbf{Lower body} & \textbf{Upper body}\\
\hline
Joint / ligament & 14 (36) & 5 (36) & 37 (43) & 14 (38)\\
\hline
Muscle / tendon & 17 (44) & 6 (43) & 45 (52) & 12 (32)\\
\hline
Other & 8 (21) & 3 (21) & 5 (6) & 11 (30)\\
\hline
\multicolumn{5}{l}{\textit{Note: }}\\
\multicolumn{5}{l}{Other included bone, skin and brain injuries.}\\
\end{tabular}
\end{table}

@@

The first network structure obtained from the data (Fig \ref{fig:fig6}) examined the interactions between explanatory variables, independent variables and probability of injury across time points in the study.
Strong sport-related and gender-based connections between several explanatory and independent variables were demonstrated for individual and team-based sports.
The ``ind\_team'' node had strong arcs to hours (0.94) and nlebase (0.78).
Individual athletes were more likely to have ``High'' hours per week (0.84) compared to team-based athletes (0.60).
Individual athletes were also more likely to have ``High'' negative life events in the 12 months preceding the start of the study compared to team-based athletes (individual athletes = 0.65, team-based athletes = 0.41).
The arc from clevel \(\rightarrow\) balance\_1 had a strength of 0.47, with lower level performers more likely to have decreased balance ability (0.48), compared to national level athletes (0.29).
High gender-based connections were reported for the arcs from gender \(\rightarrow\) stiffness\_1 (0.71) and gender \(\rightarrow\) stiffness\_2 (0.43), with males more likely to have ``High'' stiffness compared to females (males = 0.62, females = 0.43).
Irrespective of sport or gender, strong connections were found between explanatory variables. The arc from nlebase \(\rightarrow\) RI\_1 had a strength of 0.84, and the probability of RI\_1 being in the ``High'' state increased from 0.47 to 0.77 when nlebase increased from ``Low'' to ``High''.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig6-1.pdf}
\caption{\label{fig:fig6}The full Bayesian network structure was plotted using the \emph{strength.plot} function in bnlearn. The strength of each arc is shown graphically by the style of the arc. Thin, dashed arcs indicate the weakest arcs (arc strength below 0.50), whereas thick solid arcs indicate the strongest arcs (arc strength of 1). White nodes in the network indicate the explanatory variables, blue nodes indicate T1\_1 and T2\_1 variables, and red nodes indicated T2\_2 and T3\_2 variables. The injured\_X nodes have been coloured gold as they are the main nodes of interest within the network.}
\end{figure}

@@

The first conditional probability query (CPQ) investigated the variables in the Markov blanket for injured\_1 (Fig \ref{fig:fig7}), which contained hours spent training per week (hours), negative life events (nlelg\_1), muscle stiffness (stiffness\_1), current competitive level (clevel) and balance (balance\_1).
The arc between nlelg\_1 and injured\_1 was fixed in the network, so has the maximum strength of 1.

@@

{\centering \includegraphics[width=1\linewidth]{figures_doc/Fig7}

@@

\caption{Markov blanket of injured\_1. Arc strengths are included as arc labels.}\label{fig:fig7}
\end{figure}

The CPQ for injured\_1 in the ``injured'' state for all variables that were linked to injured\_1 is shown in Table \ref{tab:table5}.
The probability of injured\_1 = ``injured'' rose from 0.17 to 0.31 when stiffness was ``High'' compared to ``Low''.
Negative life events had a negligible effect when moving from the ``Low'' to ``High'' state.

@@

\caption{\label{tab:table5}Probability of injured\_1 being in the ``injured'' state, conditional on each variable}
\centering
\begin{tabular}[t]{l|r|r}
\hline
\textbf{Variable} & \textbf{Low} & \textbf{High}\\
\hline
balance\_1 & 0.21 & 0.30\\
\hline
hours & 0.18 & 0.28\\
\hline
nlelg\_1 & 0.24 & 0.26\\
\hline
stiffness\_1 & 0.17 & 0.31\\
\hline
\end{tabular}
\end{table}

The second CPQ investigated the outcome of injured\_1 being ``injured'' conditional on all variables in the Markov blanket.
The Markov blanket contained five nodes, each with two possible states resulting in \(2^5\) combinations of variables, therefore only the three lowest and highest probabilities are shown in Table \ref{tab:table6} (complete results in supplementary table 4).
The combination of lower competitive level, ``High'' hours per week, ``Low'' negative life events, ``High'' balance and ``High'' stiffness resulted in a probability of 0.53 for injured\_1 being in the ``injured'' state.
When all variables were in the ``Low'' state the probability of ``injured'' was approximately 0.04

@@

\caption{\label{tab:table6}Highest and lowest probability of injured\_1 being in the 'injured' state, conditional on the all variables in the Markov blanket for injured\_1.}
\centering
\begin{tabular}[t]{c|c|c|c|c|c}
\hline
\textbf{Probability} & \textbf{clevel} & \textbf{hours} & \textbf{nlelg\_1} & \textbf{stiffness\_1} & \textbf{balance\_1}\\
\hline
\multicolumn{6}{l}{\textbf{Highest}}\\
\hline
\hspace{1em}0.53 & club\_university\_county & High & Low & High & High\\
\hline
\hspace{1em}0.46 & national\_international & High & Low & High & Low\\
\hline
\hspace{1em}0.44 & national\_international & High & Low & High & High\\
\hline
\multicolumn{6}{l}{\textbf{Lowest}}\\
\hline
\hspace{1em}0.06 & national\_international & Low & Low & Low & Low\\
\hline
\hspace{1em}0.05 & national\_international & Low & Low & Low & High\\
\hline
\hspace{1em}0.04 & club\_university\_county & Low & Low & Low & Low\\
\hline
\end{tabular}
\end{table}

@@

The Markov blanket for injured\_2 is shown in Figure \ref{fig:fig8} and contained gender, FFFS\_1, stiffness\_2, balance\_2 and hrv\_2.
The arc between stiffness\_2 \(\rightarrow\) injured\_2 was comparable to the arc between stiffness\_1 \(\rightarrow\) injured\_1.
Very weak arcs (0.3) between injured\_2 \(\rightarrow\) balance\_2 and injured\_2 \(\rightarrow\) hrv\_2 were also present in the Markov blanket for injured\_2.
Similar to injured\_1, stiffness\_2 doubled the probability of injured\_2 being ``injured'' from 0.13 in the ``Low'' state to 0.27 in the ``High'' state.
FFFS\_1 in the ``Low'' state increased probability of injured\_2 being ``injured'' by 0.19 compared to the ``High'' state.
``High'' negative life events decreased the probability of injury from 0.26 to 0.17.

@@

{\centering \includegraphics[width=1\linewidth]{figures_doc/Fig8}

@@

\caption{Markov blanket for injured\_2.}\label{fig:fig8}
\end{figure}

>>>>

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\makeatother
\shorttitle{Stress and sports injury}
\author{Harry Fisher\textsuperscript{1}, Dr Marianne Gittoes\textsuperscript{1}, Professor Lynne Evans\textsuperscript{1}, Miss Leah Bitchell\textsuperscript{1}, Dr Richard Mullen\textsuperscript{2}, \& Dr Marco Scutari\textsuperscript{3}}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Cardiff Metropolitan University, Cardiff, United Kingdom\\\textsuperscript{2} Brunel University, London, United Kingdom\\\textsuperscript{3} Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA), Manno, Switzerland}
\authornote{12856 words in text body and refence section

Number of tables: 10

Number of figures: 5

@@

@@

Postural stability was assessed with a modified version of the balance error scoring system (mBESS) based on the protocol recommended by (55).
In total, each trial of the mBESS was performed without shoes (56) and included six stances in the following order; dominant leg (DL; standing on the dominant foot with the non-dominant foot at approximately 30-degrees of hip flexion and 45-degrees of knee flexion),
non-dominant leg (NDL; standing on the non-dominant foot with the dominant foot at approximately 30-degrees of hip flexion and 45-degrees of knee flexion)
and tandem leg stance (TS; standing heel-to-toe with the non-dominant foot behind the dominant) on firm and foam (Alcan airex AG, Sins, Switzerland) surfaces respectively (supplementary figure 2).
To determine leg dominance, participants were asked their preferred leg to kick a ball to a target, and the chosen limb was labelled as dominant (57).
Participants were asked to maintain each stance for a total of 20-seconds.
Participants hands were placed on hips at the level of the iliac crests.
A Sony DSC-RX10 video camera (Sony Europe Limited, Surrey, United Kingdom) was used to record each participants performance during the mBESS.

@@

At the start of the academic year (September), coaches of sports teams at a British University and local sports clubs were contacted and informed about the study. With the coaches' permission, the lead researcher attended training sessions to inform athletes about the overall purpose of the study and the requirements of participation. Athletes who met the participation criteria and volunteered to take part in the study were invited to attend scheduled testing sessions. A repeated measures prospective cohort design was used to assess athletes' major life events, stress-related physiological markers and injury status over two consecutive 12-month periods. Each participant was asked to attend a total of three data collections over a 12-month period, with each data collection separated by a four-month interval (supplementary figure 3). Participants provided informed consent before data collection commenced.

For the first three data collections (T1, T2 and T3), participants followed the same protocol in a specific order (supplementary figure 4).
To ensure all measures could be collected within a viable time-frame, participants were separated into two groups.
The first group completed all computer-based measures followed by all physical measurements, whereas the second group completed all physical measurements followed by computer-based measures.
Participants were randomly assigned to one of the two groups and remained in those groups across all time points.

@@

@@

Learning the structure of the network is an important step in BN modelling. The structure of a network can be constructed using expert knowledge and/or data-driven algorithm techniques (e.g., search and score, such as hill climbing and gradient descent algorithms; Scutari \& Denis, 2014).
The learned structure can then be used for inference by querying the network\footnote{The term ``query'' in relation to Bayesian Networks stems from Pearl's expert systems theory (1988). A query can be submitted to an expert (in this case, the network is the expert) to get an opinion, the expert then updates the querier's beliefs accordingly. Widely used texts on Bayesian Network analysis (71) have widely adopted the terminology in favour of that used in traditional statistics.} and obtaining the posterior probabilities of a particular node for a given query.
The posterior distribution can be obtained by \(Pr(X|E,B) = Pr(X|E,G,\Theta)\), where the learned network \(B\) with structure \(G\) and parameters \(\Theta\), are investigated with new evidence \(E\) using the information in \(B\) (38).
When a network contains many nodes, the outcome of a particular node can be assessed conditional on the states of any subset of nodes in the network.
BNs therefore provide a unique and versatile approach to modelling a set of variables to uncover dependency structures within the data.

@@

To obtain the network structure, several steps were taken to ensure that both a theoretically realistic network, and a network that was an appropriate fit to the collected data, was used for inference.
Prior knowledge about the network structure was included by providing a list of arcs that are always \emph{restricted} from being in the network (blacklist), and a list of arcs that are always \emph{included} in the network (whitelist).
Additionally, there are several scoring functions such as Bayesian Information Criteria (BIC) and Bayesian Dirichlet equivalent uniform (BDeu) that can be used to compare network structures with certain nodes and arcs included or excluded (38).
To account for the repeated measures design employed in this study and to maximise the use of the data, pairs of complete cases (e.g., participants who completed T1 + T2, and T2 + T3) were used in a two-time Bayesian network (2TBN) structure (79).
In the 2TBN, variables measured at T2 could depend on variables measured at T1 (e.g., T1 \(\rightarrow\) T2) and variables measured at T3 could depend on variables measured at T2 (e.g., T2 \(\rightarrow\) T3).
However, arcs were blacklisted between T2 \(\rightarrow\) T1 and T3 \(\rightarrow\) T2 to preserve the order in which data was collected.
Variables were separated into two groups; ``explanatory'', for variables that were fixed (e.g., gender), or ``independent'', for variables that were measured at each time point and could vary during the study.
Independent variable names were suffixed with \_1 for time point T, and \_2 for time point T+1 (e.g., T1\_1 \(\rightarrow\) T2\_2 and T2\_1 \(\rightarrow\) T3\_2).
Formatting the data in this way meant participants who completed T1 and T2, but did not complete T3, could still be included in the analysis.
In addition to the blacklisted arcs between T2 \(\rightarrow\) T1 and T3 \(\rightarrow\) T2, the direction of arcs was restricted between independent variables and explanatory variables (e.g., independent \(\rightarrow\) explanatory); however, arcs were not restricted between explanatory \(\rightarrow\) independent variables.
Finally, arc direction was restricted between specific nodes within the explanatory variables.
Arcs from clevel \(\rightarrow\) gender, nlebase \(\rightarrow\) gender and nlebase \(\rightarrow\) ind\_team were included in the blacklist, as arcs in these directions did not make logical sense.
All subsequent models used the same blacklist.

@@

The subscales representing the BAS (RR, RI, GDP and I) showed limited connection to other variables in the network.
Therefore, competing models were examined and BIC scores compared to establish the model with the best fit to the data (values are shown relative to the highest value).
RI provided the highest BIC value (-3,563.13), compared to RR (-3,579.10), GDP (-3,582.39) and I (-3,582.89).
Including all the variables (RR, RI, GDP and I) resulted in a significantly lower score -4,463.25) indicating that including all the variables was not beneficial to the model structure and did not offset the cost of the additional parameters. Therefore, only RI was included in the final structure.
Finally, both total score and asymmetry for balance were included in the initial network. However, visual inspection of the network revealed no arcs between bal\_asym\_1 or bal\_asym\_2 and any other node in the network.
Therefore, balance asymmetry was removed from the final network structure.
To summarise, Table \ref{tab:table2} reports the variables that were included in the final network structure.

@@

\caption{\label{tab:table2}Variables included in the final Bayesian network structure.}
\centering
\begin{tabular}[t]{l|>{\raggedright\arraybackslash}p{5.2cm}|c|c}
\hline
Variable & Definition & State 1 & State 2\\
\hline
clevel & Current competitive level & Club\_university\_county & National\_international\\
\hline
gender & Gender of the participant & Female & Male\\
\hline
hours & Number of hours spent training per week & 0-9 (Low) & >9-35 (High)\\
\hline
ind\_team & Participate in an individual or team based sport & Individual & Team\\
\hline
pi & Previous injury - Whether an injury had been sustained in the previous 12 months prior to the study & No Injury & Injury\\
\hline
nlebase & Untransformed NLE at TP 1 & 0-13 (Low) & >13-93 (High)\\
\hline
FFFS & Fight-Flight-Freeze System & 8-16 (Low) & >16-30 (High)\\
\hline
BIS & Behavioural Inhibition System & 17-38 (Low) & >38-68 (High)\\
\hline
RI & Reward Interest & 4-10 (Low) & >10-16 (High)\\
\hline
stiffness & Sum of all stiffness locations & 1543-2330 (Low) & >2330-4518 (High)\\
\hline
HRV & Root mean squared difference of successive RR intervals & 2.03-4.01 (Low) & >4.01-5.94 (High)\\
\hline
balance & Total balance score & 5-15 (Low) & >15-46 (High)\\
\hline
nlelg\_1 & Log NLE at TP 1 & 0-2.64 (Low) & >2.64-4.54 (High)\\
\hline
nlelg\_2 & Log NLE at TP 2 & 0-3.04 (Low) & >3.04-5.19 (High)\\
\hline
nlelg\_3 & Log NLE at TP 3 & 0-3.18 (Low) & >3.18-4.79 (High)\\
\hline
\end{tabular}
\end{table}

@@

During the study, 46\% (\emph{n} = 117) of participants reported at least one injury with an average severity of 11 \(\pm\) 31, days (range = 2 - 365 days).
Both male and female participants reported a greater number of acute compared to chronic injuries (male, acute = 85 {[}69\%{]}, chronic = 39 {[}31\%{]}; female, acute = 38 {[}72\%{]} chronic = 15 {[}28\%{]}), and non-contact injuries were more common than contact injuries (male, non-contact = 83 {[}67\%{]}, contact = 39 {[}31\%{]}; female, non-contact = 35 {[}66\%{]} contact = 18 {[}34\%{]}).
Table \ref{tab:table3} shows the number and percentage of injury types sustained by both male and female participants.

@@

\caption{\label{tab:table3}The number and percentage (\%) of types of injuries sustained by male and female participants.}
\centering
\begin{tabular}[t]{l|c|c|c|c}
\hline
\multicolumn{1}{c|}{\textbf{ }} & \multicolumn{2}{c|}{\textbf{Female}} & \multicolumn{2}{c}{\textbf{Male}} \\
\cline{2-3} \cline{4-5}
\textbf{} & \textbf{Lower body} & \textbf{Upper body} & \textbf{Lower body} & \textbf{Upper body}\\
\hline
Joint / ligament & 14 (36) & 5 (36) & 37 (43) & 14 (38)\\
\hline
Muscle / tendon & 17 (44) & 6 (43) & 45 (52) & 12 (32)\\
\hline
Other & 8 (21) & 3 (21) & 5 (6) & 11 (30)\\
\hline
\multicolumn{5}{l}{\textit{Note: }}\\
\multicolumn{5}{l}{Other included bone, skin and brain injuries.}\\
\end{tabular}
\end{table}

@@

The first network structure obtained from the data (Fig \ref{fig:fig1}) examined the interactions between explanatory variables, independent variables and probability of injury across time points in the study.
Strong sport-related and gender-based connections between several explanatory and independent variables were demonstrated for individual and team-based sports.
The ``ind\_team'' node had strong arcs to hours (0.94) and nlebase (0.78).
Individual athletes were more likely to have ``High'' hours per week (0.84) compared to team-based athletes (0.60).
Individual athletes were also more likely to have ``High'' negative life events in the 12 months preceding the start of the study compared to team-based athletes (individual athletes = 0.65, team-based athletes = 0.41).
The arc from clevel \(\rightarrow\) balance\_1 had a strength of 0.47, with lower level performers more likely to have decreased balance ability (0.48), compared to national level athletes (0.29).
High gender-based connections were reported for the arcs from gender \(\rightarrow\) stiffness\_1 (0.71) and gender \(\rightarrow\) stiffness\_2 (0.43), with males more likely to have ``High'' stiffness compared to females (males = 0.62, females = 0.43).
Irrespective of sport or gender, strong connections were found between explanatory variables. The arc from nlebase \(\rightarrow\) RI\_1 had a strength of 0.84, and the probability of RI\_1 being in the ``High'' state increased from 0.47 to 0.77 when nlebase increased from ``Low'' to ``High''.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-1.pdf}
\caption{\label{fig:fig1}The full Bayesian network structure was plotted using the \emph{strength.plot} function in bnlearn. The strength of each arc is shown graphically by the style of the arc. Thin, dashed arcs indicate the weakest arcs (arc strength below 0.50), whereas thick solid arcs indicate the strongest arcs (arc strength of 1). White nodes in the network indicate the explanatory variables, blue nodes indicate T1\_1 and T2\_1 variables, and red nodes indicated T2\_2 and T3\_2 variables. The injured\_X nodes have been coloured gold as they are the main nodes of interest within the network.}
\end{figure}

@@

The first conditional probability query (CPQ) investigated the variables in the Markov blanket for injured\_1 (Fig \ref{fig:fig2}), which contained hours spent training per week (hours), negative life events (nlelg\_1), muscle stiffness (stiffness\_1), current competitive level (clevel) and balance (balance\_1).
The arc between nlelg\_1 and injured\_1 was fixed in the network, so has the maximum strength of 1.

@@

@@

\caption{Markov blanket of injured\_1. Arc strengths are included as arc labels.}\label{fig:fig2}
\end{figure}

The CPQ for injured\_1 in the ``injured'' state for all variables that were linked to injured\_1 is shown in Table \ref{tab:table4}.
The probability of injured\_1 = ``injured'' rose from 0.17 to 0.31 when stiffness was ``High'' compared to ``Low''.
Negative life events had a negligible effect when moving from the ``Low'' to ``High'' state.

@@

\caption{\label{tab:table4}Probability of injured\_1 being in the ``injured'' state, conditional on each variable}
\centering
\begin{tabular}[t]{l|r|r}
\hline
\textbf{Variable} & \textbf{Low} & \textbf{High}\\
\hline
balance\_1 & 0.21 & 0.30\\
\hline
hours & 0.18 & 0.28\\
\hline
nlelg\_1 & 0.24 & 0.26\\
\hline
stiffness\_1 & 0.17 & 0.31\\
\hline
\end{tabular}
\end{table}

The second CPQ investigated the outcome of injured\_1 being ``injured'' conditional on all variables in the Markov blanket.
The Markov blanket contained five nodes, each with two possible states resulting in \(2^5\) combinations of variables, therefore only the three lowest and highest probabilities are shown in Table \ref{tab:table5} (complete results in supplementary table 4).
The combination of lower competitive level, ``High'' hours per week, ``Low'' negative life events, ``High'' balance and ``High'' stiffness resulted in a probability of 0.53 for injured\_1 being in the ``injured'' state.
When all variables were in the ``Low'' state the probability of ``injured'' was approximately 0.04

@@

\caption{\label{tab:table5}Highest and lowest probability of injured\_1 being in the ``injured'' state, conditional on the all variables in the Markov blanket for injured\_1.}
\centering
\begin{tabular}[t]{c|c|c|c|c|c}
\hline
\textbf{Probability} & \textbf{clevel} & \textbf{hours} & \textbf{nlelg\_1} & \textbf{stiffness\_1} & \textbf{balance\_1}\\
\hline
\multicolumn{6}{l}{\textbf{Highest}}\\
\hline
\hspace{1em}0.53 & club\_university\_county & High & Low & High & High\\
\hline
\hspace{1em}0.46 & national\_international & High & Low & High & Low\\
\hline
\hspace{1em}0.44 & national\_international & High & Low & High & High\\
\hline
\multicolumn{6}{l}{\textbf{Lowest}}\\
\hline
\hspace{1em}0.06 & national\_international & Low & Low & Low & Low\\
\hline
\hspace{1em}0.05 & national\_international & Low & Low & Low & High\\
\hline
\hspace{1em}0.04 & club\_university\_county & Low & Low & Low & Low\\
\hline
\end{tabular}
\end{table}

@@

The Markov blanket for injured\_2 is shown in Figure \ref{fig:fig3} and contained gender, FFFS\_1, stiffness\_2, balance\_2 and hrv\_2 (table \ref{tab:table6}).
The arc between stiffness\_2 \(\rightarrow\) injured\_2 was comparable to the arc between stiffness\_1 \(\rightarrow\) injured\_1.
Very weak arcs (0.3) between injured\_2 \(\rightarrow\) balance\_2 and injured\_2 \(\rightarrow\) hrv\_2 were also present in the Markov blanket for injured\_2.
Similar to injured\_1, stiffness\_2 doubled the probability of injured\_2 being ``injured'' from 0.13 in the ``Low'' state to 0.27 in the ``High'' state.
FFFS\_1 in the ``Low'' state increased probability of injured\_2 being ``injured'' by 0.19 compared to the ``High'' state.
``High'' negative life events decreased the probability of injury from 0.26 to 0.17.

@@

{\centering \includegraphics[width=1\linewidth]{figures_doc/Fig3}

@@

\caption{Markov blanket for injured\_2.}\label{fig:fig3}
\end{figure}
